---
title: "Git Notes again"
format: html
---

## Basics of using git 

Git is a distruted version control system or VCS that is very popular. 

The core steps to using git involve

1. Initiate a git repository 
  In RStudio we just click a little radio button on project creation at the command line we type:
  `git init`
  
2. Do your normal work 
  Add files
  Modify files, etc. 
  
3. Add your important files to the "stage"
  In Rstudio we click "staged" radio button
  `git add yourfilename`

4. Commit your staged files to the repository
  In RStudio you just hit the "Commit" button and add a msg
  `git commit -m "your wee message"`
  

You can check your project "history" and status
  `git status`
  `git log` ( push Q to exit )
  
5. GIT PUSH TO SYNC CHANGES TO GITHUB
  The Up arrow in Rstudio`git push`
  
6. Git pull to sync changes to github
  The up arrow in RStudio

  


## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this 

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
