---
title: "Class 08 : Breast Cancer Mini Project"
author: "Kamila Medina PID: A18475433"
format: pdf
toc: true
---

## Background 

In today's class we will be employing all R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy data. 

## Data Import 

The data is in CSV format:

```{r}
wisc.df <- read.csv ("WisconsinCancer.csv", row.names=1)
```

peek at the data 
```{r}
head(wisc.df, 3)
```
> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many of the observations have a malignant diagnosis?


```{r}
sum(wisc.df$diagnosis == "M")
```
```{r}
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean",colnames(wisc.df)))
```

```{r}
colnames(wisc.df)
```

We need to removed the `diagnosis` column before we do any further analysis of this dataset - we don't want to pass this to PCA etc. We will save it as a separate vector that we can use later to compare our findings to those of experts. 

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Prinicpal Component Analysis (PCA)

The main function in base R is called `prcomp()` we will use the optional argument `scale=TRUE` here as the data columns/features/ dimensions are very different scales in the original data set. 

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
```

```{r}
attributes(wisc.pr)
```
```{r}
wisc.pr$x
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x) + 
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```
```{r}
summary(wisc.pr)
```

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", "PC1"]
```


## Hierarchical Clustering 

The goal of this section is to do hierarchical clustering of the **original data** to  see if there is any obvious grouping into malignant and benign clusters.


```{r}
wisc.dist <- dist(scale(wisc.data))
wisc.hclust <- hclust(wisc.dist)
plot(wisc.hclust)
```
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=2)
table(wisc.hclust.clusters)
```
## Combining Methods

 The idea here is that I can take my new variables (i.e. the sores on PCs `wisc.pr$x) that are better descriptors of the data-set than the original features (i.e. the 30 columns in `wisc.data`) and use these as a basis for clustering.

```{r}
pc.dist <- dist(wisc.pr$x[ ,1:3])
wisc.pr.hclust <- hclust(pc.dist, method ="ward.D2")
plot(wisc.pr.hclust)
```
```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(diagnosis)
```

I can now run `table()` with both my clustering `grps` and the expert `diagnosis`
```{r}
table(grps, diagnosis)
```

Our cluster "1" has 179 "M" diagnosis 
our cluster "2" has 333 "B" diagnosis

179 TP
24 FP
333 TN
33 FN 

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity : TN/(TN+FP)
```{r}
333/(333+24)
```
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

