---
title: "Class 09 : Candy Mini Project"
author: "Kamila Medina PID A18475433"
format: pdf
toc: true 
---
## Importing Candy Data 
```{r}
candy_file <- "candy-data.csv"
candy <- read.csv(candy_file,stringsAsFactors = FALSE)
```

```{r}
candy = read.csv(candy_file, row.names = 1)
head(candy)
str(candy_file)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
## Winpercent

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Almond Joy", ]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

## Skim 
```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes. The sugarpercent for example. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
skim(candy$chocolate)
```
In the chocolate column, the 1 represents the candy contains chocolate and the 0 the candy does not contain chocolate. This is a **classic binary indicator** and skim() function shows it as having only zeros and ones.

## Exploratory Analysis 

> Q8. Plot a histogram of winpercent values using both base R an ggplot2

```{r}
hist(candy$winpercent,
     main = "Histogram of Candy Win Percent",
     xlab = "Win Percent",
     col = "lightpink",
     breaks = 10)

library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(binwidth = 5, fill = "lightpink", color = "black") +
  labs(title = "Histogram of Candy Win Percent",
       x = "Win Percent",
       y = "Count")
```


> Q9. Is the distribution of winpercent values symmetrical?

No, it's slightly right skewed. 

> Q10. Is the center of the distribution above or below 50%?

It's above 50%. 


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate <- candy$winpercent[as.logical(candy$chocolate)]
fruity <- candy$winpercent[as.logical(candy$fruity)]

mean(chocolate)
mean(fruity)
```

```{r}
choc.candy <- candy[candy$chocolate==1,]
choc.win <- choc.candy$winpercent
mean(choc.win)
```
```{r}
fruity.candy <- candy[candy$fruity==1,]
fruity.win <- fruity.candy$winpercent
mean(fruity.win)
```

Chocolate candy is ranked higher. 


> Q12. Is this difference statistically significant?

```{r}
chocolate <- candy$winpercent[as.logical(candy$chocolate)]
fruity <- candy$winpercent[as.logical(candy$fruity)]

t.test(chocolate, fruity,var.equal = TRUE)
```


```{r}
t.test(choc.win, fruity.win)
```

The difference is statistically significant because the p-value is far below 0.05 and the confidence interval does not include 0. 

## Overall Candy Rankings 
> Q13. What are the five least liked candy types in this set?

```{r}
sort(candy$winpercent)
```
```{r}
inds <- order(candy$winpercent)
head(candy[inds,],5)
```


```{r}
head(candy[order(candy$winpercent),], n=5)
```
```{r}
candy |> arrange(winpercent) |> head(5)
```
```{r}
head(candy[order(candy$winpercent), ], 5)
```
 
I prefer the base R function over dplyr. 

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}

head(candy[order(-candy$winpercent), ], 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()

```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() +
  ylab("") # turn off the y label we don't need 

ggsave("barplot1.png", height=10, width=6)
```
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "pink"
my_cols[as.logical(candy$fruity)] = "lightgreen"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
> Q17. What is the worst ranked chocolate candy?

```{r}
chocolate_candies <- candy[as.logical(candy$chocolate), ]
rownames(chocolate_candies)[which.min(chocolate_candies$winpercent)]

```


> Q18. What is the best ranked fruity candy?

```{r}
fruity_candies <- candy[as.logical(candy$fruity), ]
rownames(fruity_candies)[which.max(fruity_candies$winpercent)]
```

## PricePercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x = pricepercent, y = winpercent, label = rownames(candy)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 5)

```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


```{r}
value_score <- candy$winpercent / candy$pricepercent
rownames(candy)[which.max(value_score)]
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular? 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)

top5_expensive <- candy[ord, ][1:5, c("pricepercent", "winpercent")]
top5_expensive
```
```{r}
rownames(top5_expensive)[which.min(top5_expensive$winpercent)]
```
> Q21. Make a barplot again using pricepercent and then improve this

```{r}
ggplot(candy) +
  aes(x = reorder(rownames(candy), pricepercent), y = pricepercent) +
  geom_segment(aes(xend = reorder(rownames(candy), pricepercent), y = 0, yend = pricepercent), color = "gray") +
  geom_point(color = "skyblue", size = 3) +
  coord_flip() +
  labs(x = "Candy", y = "Price Percent", title = "Candy Price Percent Lollipop Chart")

```
```{r}
ggplot(candy) +
  aes(x = pricepercent, y = reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(x = 0, xend = pricepercent, 
                   y = reorder(rownames(candy), pricepercent), 
                   yend = reorder(rownames(candy), pricepercent)),
               color = "gray") +
  geom_point(color = "lightpink", size = 3) +
  labs(x = "Price Percent", y = "Candy", title = "Candy Price Percent Lollipop Chart") +
  theme_minimal()
```

## Exploring Correlation Structure
Pearson correlation values range from -1 to +1
```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity 

> Q23. Similarly, what two variables are most positively correlated?

chocolate and peanutyalmondy

## Principal Component Analysis 

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)

```

the main results figure: the PCA score plot:
```{r}
ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point(col=my_cols) + 
  labs(title="PCA CANDY SPACE MAP", 
       subtitle= "Separation of choco")
```

```{r}
plot(pca$x[,1], pca$x[,2],
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA Score Plot: PC1 vs PC2",
     pch = 19, col = "lightpink")

```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
library(ggplot2)
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(cbind(candy, pca$x[,1:3])) +
        aes(x = PC1, y = PC2,
            size = winpercent / 100,
            text = rownames(candy),
            label = rownames(candy)) +
        geom_point(col = my_cols)
p
```

we can use **ggrepel** to improve the plot. 

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 3)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data")
```


```{r}
library(ggplot2)

loadings <- as.data.frame(pca$rotation)
loadings$variable <- rownames(loadings)

ggplot(loadings) +
  aes(x = reorder(variable, PC1), y = PC1, fill = PC1 > 0) +
  geom_col() +
  coord_flip() +  # horizontal bars
  labs(title = "PCA Loadings for PC1",
       x = "Variable",
       y = "PC1 Loading") +
  scale_fill_manual(values = c("lightpink", "skyblue"), guide = FALSE) +
  theme_minimal()


```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?
 
Winpercent, chocolate and bar were picked up. They do make sense, it was highlighted in the correlation plot and t tests.


> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A winning candy seems to be a chocolate + bar + peanuty combo, and the  fruity candies seem less likely to be “winners.” The PCA and other graphs back this up. 

> Q26. Are popular candies more expensive?

```{r}

losers <- candy[which(candy$winpercent < 50), ]
winners <- candy[which(candy$winpercent >= 50), ]

mean(losers$pricepercent)
mean(winners$pricepercent)

t.test(winners$pricepercent, losers$pricepercent)

```
Not really, chocolate bars kind of are. 

> Q27. Are candies with more sugar more likely to be popular? What is your interpretation of the means and P-value in this case?

```{r}

mean(losers$sugarpercent)
mean(winners$sugarpercent)

t.test(winners$sugarpercent, losers$sugarpercent)

```
 Kind of correlated. 



