---
title: "Class06: R Functions"
author: "Kamila Medina PID: A18475433"
format: pdf
toc: true 
---

## Background

Functions are at the heart of using R. Everything we do involves calling and using functions (from the data input, analysis to results output).


All functions in R have at least 3 things:

1. A **name** the thing we use to call the function.
2. One or more input **arguments** that are comma separated
3. The **body**, lines of code between curly brackets { } that does the work of the function. 

## A First Function

Let's write a silly little function to add some numbers:
```{r}
add <- function(x) {x + 1}
```

Let's try it out 
```{r}
add(100)
```
Will this work
```{r}
add( c(100,200,300))
```
Modify to be more useful and add more than just 1
```{r}
add <- function(x,y=1) {x + y}
```

```{r}
add(100,10)
```
```{r}
add(100)
```
```{r}
plot(1:10, col="pink", typ="b")
```
```{r}
log(10, base=10)
```
> **N.B.** Input arguments can be either **required** or **optional**. The latter have a fall-back default that is specified in the function code with an = sign.  

```{r}
#add(x=100,y=200,z=300)
```

## A Second Function

All functions in R look like this 

```
name <- function(arg) {body}
```

The 'sample()` function in R...
```{r}
sample(1:10, size = 4)
```
> Q. Return 12 numbers randomly from the input 1:10

```{r}
sample(1:10, size = 12, replace = TRUE)
```
> Q. Write the code to generate a random 12 nuceotide long DNA sequence?

```{r}
bases <-c("A","T","G", "C")
sample(bases, size = 12, replace = TRUE)
```
> Q. Write a first verse function called `generate_dna()` that generates a user specified length random DNA sequence? 

```{r}
generate_dna <- function(n=6) {
bases <- c("A","T","G", "C")
sample(bases, size=n, replace=TRUE) }
```
```{r}
generate_dna(100)
```
> Q. Modify your function to return a FASTA like sequence so rather than [1] "A" "G" "T" "C" we want "AGTC"

```{r}
generate_dna <- function(n=6) {
bases <- c("A","T","G", "C")
ans <- sample(bases, size=n, replace=TRUE) 
paste(ans, collapse = "")}
```
```{r}
generate_dna(10)
```
> Q. Give the user an option to return FASTA format output sequence or standard multi element vector format? 

```{r}
generate_dna <- function(n=6, fasta=TRUE) {
bases <- c("A","T","G", "C")
ans <- sample(bases, size=n, replace=TRUE)

if(fasta) {
paste(ans, collapse = "")}}
```
```{r}
generate_dna(10)
generate_dna(10, fasta=F)
```
## A new cool function 

> Q. Write a function called `generate_protein()` that generates a user specified length protein sequence in FASTA like format? 

```{r}
generate_protein <- function(n) {
aa <- c("A", "R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")

ans <- sample (aa, size = n, replace = T)
paste(ans, collapse = "")}
```
```{r}
generate_protein(12)
```



> Q. Use your new `protein_sequence()` function to generate sequences between length 6 and 12 amino acids in length and check if any of these are unique in nature (i.e. found in the NR database at NCBI)?

```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
generate_protein(10)
generate_protein(11)
generate_protein(12)
```

Or we could do a `for()` loop:

```{r}
for (i in 6:12) {
  cat(">", i, sep="", "\n")
  cat( generate_protein(i), "\n" )
}
```
Some are and some are not unique. 





