---
title: "Class 7 : Machine Learning 1"
author: "Kamila Medina PID: A18475433"
format: pdf
---

## Background
Today we will begin our exploration of some important machine learning methods, namely **clustering** and **dimensionality reduction**. 

Let's make up some input data for clustering where we know what the natural "clusters" are. 

The function `rnorm()` can be useful here. 

```{r}
hist(rnorm(5000))
```
>Q. Generate 30 random numbers centered at +3 and another 30 centered at -3

```{r}
tmp <- c(rnorm(30, mean=3),
rnorm(30, mean=-3))

x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

## K-means clustering

The main function in "base R" for k-means is called `kmeans()`:

```{r}
km <- kmeans(x, centers = 2)
```
> Q. What component of the results object details the cluster sizes? 

```{r}
km$size
```

> Q. What component of the results object details the cluster centers? 

```{r}
km$centers
```


> Q. What component of the result object details the cluster membership vector?

```{r}
km$cluster
```

> Q. Plot our clustering results with points colored by cluster and also add the cluster centers as new points colored blue. 

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15)
```
> Q. Run `kmeans()` again and this time produce 4 clusters and call your result object `k4` and make a results figure like above? 

```{r}
set.seed(123)
k4 <- kmeans(x, centers = 4)

plot(x, col=k4$cluster)
points(k4$centers, col="blue", pch=15)

k4 <- kmeans(x, centers = 4) 

```

The metric
```{r}
km$tot.withinss
k4$tot.withinss
```
> Q. Let's try different numbers of K (centers) from 1 to 30 and see what's the best result? 

```{r}
i <- 1
ans <- NULL
for(i in 1:30){
ans <- c(ans, kmeans(x,centers = i)$tot.withinss)
}

ans
```

```{r}
plot(ans,typ="o")
```
**Note to self** : K-means is ggonna make a cluster structure on the data even if it's not there, it goves you hwat you ask for even if it doesn't really make sesne for the data 

## Hierarchical Clustering 

The main function for Hierarchical Clustering is called `hclust()`. Unlike `kmeans()` (which does the work for you) you can't just pass `hclust()` our raw input data. It needs a "distance matrix" like the one returned from the `dist()` function. 

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

To extract our cluster membership vector from a `hclust()` result object we have to "cut" our tree at a given height to yield seperate "groups"/"branches". 

```{r}
plot(hc)
abline(h=8, col="red", lty=2)
```

To do this we use the `cutree` function on our `hclust` object:

```{r}
cutree(hc, h=8)
```
```{r}
grps <- cutree(hc, h=8)
grps
```

```{r}
table(grps, km$cluster)
```

## PCA of UK food data 

Import the dataser on food consumption in the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```
> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

One solution to set the row names is to do it by hand...
```{r}
rownames(x) <- x[,1]
```

To remove the first column I can use the minus index trick
```{r}
x <- x[,-1]
```

A better way to do this is to set the row names to the first column by arguing with `read.csv`

```{r}
x <- read.csv(url, row.names = 1)
x
```
 > Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
 
 The second way is a lot better. 
 
## Spotting major differences and trends 
is diffciult with this 17D dataset...

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```
```{r}
rainbow(40)
```
### Pairs plots and heatmaps
 
```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)

```

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

### PCA to the Rescue

The main PCA function in the "base R" is called `prcomp()`. This function wants the transpose of our food data as input (i.e. the foods as columns and the countries as rows). 

```{r}
pca <- prcomp(t(x))
```

```{r}
summary(pca)
```
```{r}
attributes(pca)
```

To make one of the main PCA result figures we turn to `pca$x` the scores along our new PCs. This is called "PC plot" or score plot ot "Ordination plot"...

```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
```

```{r}
library(ggplot2)

ggplot(pca$x) + 
  aes(PC1,PC2) + 
  geom_point(col=my_cols)
```

The second major result figure is called a "loadings plot" of "variable contributions plot" or "weight plot". 

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation))+
  geom_col()
```


